HO unification from object language to meta language

### 2024-03-28 paper/main.tex
rewrite handling eta

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}

$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

@@

$\eta$-reduction is an equivalence relation over terms where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expanded
version of $t$. The implementation of the \elpiIn{comp} relation given in
\cref{sec:compilation} compiles the \Fo{} terms $t_1 = $\elpiIn{flam x\ fapp
[fuva A, x]} and $t_2 =$ \elpiIn{fcon "f"} into the \Ho terms $t'_1 =
$\elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
\mapping{A}{A'}{1}. However, if the oracle sets \elpiIn{A'} to the constant
\elpiIn{"f"}, the unification of $t'_1$ and $t'_2$ in the meta language will
fail even though $t_1 \Eo t_2$. The reason of this failure is attributed to the
fact that $t'_1 =$\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
=$\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
unification problem by adapting the \elpiIn{comp} relation such that it
recognizes \maybeeta subterms $s$ and replaces them with fresh \Ho variables
$v$. This link between the variable $v$ and the subterm $t$ is stored in what we
call \linketa which is an object with the following type

@@

Compiling term with \maybeeta terms forces us to determine if, $\lambda x. T_x$,
for any term $T$ having $x$ in scope, can be a $\eta$-expansion, i.e. under a
given substitution $\sigma$, we have $\sigma (\lambda x.T_x) = t$. This
\maybeeta detection is not a trivial operation as it may seems.
%
\begin{gather}
  \lambda x. f \appsep A_x\\
  \lambda x. f \appsep x \appsep A_x\\
  \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}
\end{gather}

In the examples above, the first expression is a \maybeeta since $A_x$ can
reduce to $x$, the second one is not a \maybeeta since for any substitution for
$A_x$, $x$ is not free in $f \appsep x$. The third equation is a bit more
complicated since, we have a spine of lambdas, this means that the whole term
is a \maybeeta, if the inner $\lambda$-term is an $\eta$-expansion of a term
$t$, and $t$ can be reduced to a term on the form $t' x$ where $x$ is not free
in $t'$. Indeed, eq.~\ref{eq:lamlam} is a \maybeeta under the substitution 
$\sigma = \{A \mapsto \lambda x.x, B \mapsto \lambda x. \lambda y. x\}$.

TODO: clarify this
As a remark, note that $\lambda x. f \appsep A_x \appsep x$ is a \maybeeta,
since, despite, $x$ occurs in $f \appsep A_x$, it is still possible that this
subterm does not use $x$, for example if $A$ is a function on the form $\lambda
x.a$, where $a$ is a defined constant. In this case, the \maybeeta should
consider that the bound variable \elpiIn{x} does not ``rigidly'' occur in 
the given subterm.

We can now define more formally the two auxiliary relation we need for \maybeeta
detection:

@@

If we take back the example give at the beginning of this section, we can
deduplicate $\mapping{F}{G}{1}, \mapping{F}{H}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{G_{x}}{\lambda y.H_{x~y}}.

>>>>

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\newcommand{\linkStore}{\ensuremath{\mathbb{L}}\xspace}
\newcommand{\mapStore}{\ensuremath{\mathbb{M}}\xspace}
%
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

@@

$\eta$-reduction is an equivalence relation over terms where a term of the form
$\lambda x.t \appsep x$ can be converted to $t$ any time $x$ does not occur as a
free variable in $t$. We call $\lambda x. t \appsep x$ the $\eta$-expanded
version of $t$. The implementation of the \elpiIn{comp} relation given in
\cref{sec:compilation} compiles the \Fo{} terms $t_1 =$ \elpiIn{flam x\ fapp
[fuva A, x]} and $t_2 =$ \elpiIn{fcon "f"} into the \Ho terms $t'_1 =
$ \elpiIn{lam x\ uva A' x} and $t'_2 =$ \elpiIn{fcon "f"} with mapping
\mapping{\elpiIn{A}}{\elpiIn{A'}}{1}. However, if the oracle sets \elpiIn{A'} to the constant
\elpiIn{"f"}, the unification of $t'_1$ and $t'_2$ in the meta language will
fail even though $t_1 \Eo t_2$. The reason of this failure is attributed to the
fact that $t'_1 =$~\elpiIn{lam x\ app[con "f", x]} cannot be unified with $t'_2
=$~\elpiIn{con "f"} since the two terms have different rigid heads. We solve this
unification problem by adapting the \elpiIn{comp} relation such that it
recognizes \maybeeta subterms and replaces them with fresh \Ho variables. 
This link between the variable $v$ and the subterm $t$ is stored in what we
call \linketa which is an object with the following type

@@

Compiling term with \maybeeta terms forces us to determine if, $\lambda x. T_x$,
for any term $T$ having $x$ in scope, can be a $\eta$-expansion, i.e. under a
given substitution $\sigma$, we have $\sigma (\lambda x.T_x) = t$. This
\maybeeta detection is not a trivial operation as it may seems.
\\
\noindent
\begin{minipage}{.5\linewidth}
  \begin{gather}
    \lambda x. f \appsep A_x\\
    \lambda x. f \appsep x \appsep A_x
  \end{gather}
\end{minipage}% This must go next to `\end{minipage}`
\begin{minipage}{.5\linewidth}
  \begin{gather}
    \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
    \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
  \end{gather}
\end{minipage}

In the examples above, the first expression is a \maybeeta since $A_x$ can
reduce to $x$, the second one is not a \maybeeta since it exists no substitution
for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
Eq.~\ref{eq:lamlam} is a bit more
complicated since, we have a spine of lambdas, this means that the whole term
is a \maybeeta, if the inner $\lambda$-term is an $\eta$-expansion of a term
$t$, and $t$ can be reduced to a term on the form $t' x$ where $x$ is not free
in $t'$. Indeed, eq.~\ref{eq:lamlam} is a \maybeeta under the substitution 
$\sigma = \{A \mapsto \lambda x.x, B \mapsto \lambda x. \lambda y. x\}$.

As a final remark, note that eq.~\ref{eq:lamdisc} is a \maybeeta even if $x$
occurs in the subterm $t' = f \appsep A_x$. It exists in fact a substitution
$\sigma = \{A \mapsto \lambda x.a\}$, for any constant $a$, such that $\sigma t'
= f \appsep a$ is a term where $x$ does not appears as a free variable.

We can now define more formally the two auxiliary relations needed for \maybeeta
detection:

@@

If we take back the example give at the beginning of this section, we can
deduplicate $\mapping{F}{G}{1}, \mapping{F}{H}{2}$ by removing the first mapping
and adding the auxiliary \linketa: \linketaM{x}{G_{x}}{\lambda y.H_{x y}}.
